Best Practices for Multiple Alchemical Free Energy Calculations [Article v\versionnumber]

### 2025-05-26 paper/manuscript.tex
tweaks

\section{\tocorange{Large-Scale Simulation Planning}}
\label{sec:large_scale_planning}
\tocsectioncomment{Possibly add discussion of methods which are not easily characterised as single/ multi e.g. lambda dynamics? Maybe a better distinction is calculations where we are interested in more than two end states.}
\tocsectioncomment{@OpenFE team to add discussion of how to run and adaptively update perturbation networks.}
\tocsectioncomment{Possibly add discussion of different protocols, especially active learning. Best practices may not be well established but many companies are running at scale.}
\tocsectioncomment{Discussion of merits of ABFE/ RBFE for virtual screening? See @Gary Tresadern comment from first meeting - lots of discussion around ABFE calculations for virtual screening.}
{\color{blue}
When dealing with large datasets, methods that compute multiple ligands' binding free energies simultaneously offer a compelling alternative to pairwise calculations. These approaches, categorized as "multiple end state" or "multistate" free energy calculations, can be more convenient than standard methods as they sidestep ligand network construction entirely and can deliver all relative free energies in a single run. However, requiring many end states to be sampled at once complicates the energy landscape, forcing longer, inherently sequential simulations. Additionally, they are often more difficult to set up; combining multiple ligand topologies raises extra challenges about avoiding bias from distance restraints or hybrid models while requiring sufficient phase-space overlap.

A well-established method from this category is enveloping distribution sampling \cite{christ2007enveloping} (EDS), which constructs a single reference potential that samples the end state phase spaces of any number of states. Multiple EDS potentials with different levels of barrier-smoothing can be combined in a replica-exchange setup in order to promote crossing of alchemical energy barriers that separate the end-states \cite{sidler2017efficient, champion2023accelerating}. This approach (RE-EDS) confers several practical advantages. Because the protein and solvent degrees of freedom are propagated only once per time step, the computational cost of adding more ligands grows only linearly rather than quadratically (in a fully connected perturbation graph). Furthermore, RE-EDS inherits intrinsic cycle closure from the enveloping reference state, eliminating the need to check or enforce thermodynamic cycles by hand.

Setting up RE-EDS simulations requires some thought to be put in system preparation. In the dual-topology setup all ligands must be restrained to avoid diffusing out of the binding site and ensuring phase-space overlap so that intermediate configurations smoothly connect ligand energy wells. However the number and definition of restraints should be chosen carefully as they can introduce bias in the simultations if not explicitely corrected. Multistate hybrid topologies can alleviate this issue as well as improve phase-space overlap \cite{champion2023accelerating} but their implementation in most MD engines usually requires custom routines, limiting their applicability.

The RE-EDS workflow as implemented in the GROMOS \cite{champion2023accelerating} or OpenMM \cite{rieder2022leveraging} software packages includes an automated calibration procedure in which the EDS potentials on the replica-exchange ladder are iteratively optimized by minimizing round-trip times and balancing state occupancy. Nevertheless, heterogeneous datasets containing ligands that require insertion or deletion of many heavy atoms can drastically reduce the efficiency of the method, and in these cases it is generally beneficial to split very large or poorly overlapping ligand sets into smaller groups or subsets to be executed with pairwise calculations.
}

\paragraph{Perturbation maps}
\tocsectioncomment{Perturbation maps paragraph: @Mary Pitman/ @Jenke Scheen/ OpenFE team to update discussion of perturbation maps?}
Based on the input ligand series, a perturbation map or network can be planned. Recent heuristics have shown the more connected the perturbation network the better. However, there is a way to optimize network structure while minimizing the number of perturbations that need to be computed reducing the resulting computational cost~\cite{yang2020optimal,xu2019optimal}. Sometimes the introduction of intermediates that are not part of the original congeneric series are essential to avoid ring breaking, or to deal with perturbations that would otherwise result in large numbers of atoms being inserted or deleted. Some commercial tools have good underlying heuristics but may fail with complicated input, needing user validation in particular when dealing with chiral compounds.

>>>>

\section{\tocorange{Large-Scale Simulation Planning}}
\label{sec:large_scale_planning}
\tocsectioncomment{Possibly add discussion of methods which are not easily characterised as single/ multi e.g. lambda dynamics? Maybe a better distinction is calculations where we are interested in more than two end states.}
\tocsectioncomment{@OpenFE team to add discussion of how to run and adaptively update perturbation networks.}
\tocsectioncomment{Possibly add discussion of different protocols, especially active learning. Best practices may not be well established but many companies are running at scale.}
\tocsectioncomment{Discussion of merits of ABFE/ RBFE for virtual screening? See @Gary Tresadern comment from first meeting - lots of discussion around ABFE calculations for virtual screening.}

\paragraph{Perturbation maps}
\tocsectioncomment{Perturbation maps paragraph: @Mary Pitman/ @Jenke Scheen/ OpenFE team to update discussion of perturbation maps?}
For RBFE campaigns with more than two ligands in the input series a \textit{perturbation map} or \textit{-network} must be generated: this is simply a set of data that describe which transformations will be executed between the ligands (referred to as \textit{edges}). \cite{MOBLEY LOMAP} Heuristics have shown the more connected the perturbation network the higher the free energy prediction accuracy versus experimental measures. Running a fully-connected network (i.e. simulate transformations between \textit{all} ligands in the campaign) is typically considered too computationally expensive and may not be suited optimally with current estimators such as MLE (\ref{SECTION MLE}).\cite{SCHEEN ET AL} Several styles of network designs have been proposed in recent years, although no single method seems to consistently outperform all others.\cite{KARTOGRAF} There exist several methods for optimization of network structure while minimizing the number of perturbations that need to be computed, reducing the resulting computational cost at a minimal decrease in accuracy ~\cite{yang2020optimal,xu2019optimal}. Sometimes - depending on the RBFE implementation - the introduction of intermediate ligands that are not part of the original congeneric series can be helpful for avoiding challenging transformations such as ring breaking or transformations that would otherwise result in large numbers of atoms being inserted or deleted.\cite{LINDE PAPER} Some commercial tools have well-calibrated underlying heuristics to plan these network designs but may fail with complicated input datasets, thus requiring manual user validation, in particular when dealing with chiral compounds.
